Homicide Increases in the COVID-19 Era

### 2025-02-28 1_FirstDraft/Draft1.qmd
Organization

author: "Marco Stine, Kevin Hu, Bianca Schutz, and Kaitlyn Kirt"
fontsize: 10pt
toc: true                   # add table of contents at the beginning
toc-depth: 3               # Only titles that start with # or ##
geometry: margin=1in
bibliography: references.bib
format:
  pdf: default
number-sections: true
---

@@

```{r DATA, output=FALSE, echo=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses

# primary datasets
# Primary datasets: Crime Data 2020-Present and 2010-2019
crime_la_2010_2019 <- read.csv("Raw_Data/Crime_Data_from_2010_to_2019.csv")

@@

crime_la <- read.csv("Cleaned_Data/all_crimes_2010_present.csv")

# primary dataset zip code and geometry data
crimes_with_zips <- read.csv("Cleaned_Data/crimes_with_zips.csv") # just inner_join with the crime_la dataset

# Secondary dataset - Median Income by ZCTA
cali_median_income <- read.csv("Cleaned_Data/cali_median_income.csv")

# adding a variable that's just the ZCTA zip code
cali_median_income$ZCTA <- stringr::str_sub(cali_median_income$NAME, -5)

# Secondary dataset - COVID cases in Los Angeles County
covid_14day <- read.csv("Raw_Data/LA_County_Covid19_CSA_14day_case_death_table.csv")

evictions_la <- read.csv("Raw_Data/LA_evictions.csv")
```

# Exploratory Data Analysis

@@

```{r, echo=FALSE}

@@

```{r, echo=FALSE}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

@@

abline(h = pretty(weekly_counts$Crimes), v = pretty(weekly_counts$Index), 
       col = "gray", lty = "dotted")
```
### Interpretation

@@

```{r, echo=FALSE, fig.cap = "Crime Occurrences by Day of the Week"}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

# Plot bar chart with horizontal labels
barplot(crime_counts, col = "lightblue",
        xlab = "Day of the Week", ylab = "Number of Crimes",
        names.arg = names(crime_counts), las = 1, cex.names = 0.8, cex.axis = 0.6)
```

@@

```{r, echo=FALSE}
# Remove missing values and empty strings in Vict.Sex column
crime_la <- subset(crime_la, Vict.Sex != "" & !is.na(Vict.Sex))

@@

```{r, echo=FALSE}
zcta <- read_sf(dsn = "Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
LA_zctas <- read.csv("Raw_Data/zctas.csv") #obtained using zipcodeR::search_city("Los Angeles", "CA") and then checking which zip codes are also ZCTAs

@@

```{r, echo=FALSE}
# quantiles for the entire dataset (from 2018 to 2023) to divide the median income into groups
qntls <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

@@

The maps demonstrate that median income in certain LA zip codes (particularly those to the west) have a much higher median household income than the rest of the city, but over the last six years, the median income in most other LA zip codes has increased. However, there are some that a typical family is at or below the poverty line. We intend to investigate if there is a spatial correlation between these poorer zip codes and the incidence of crimes, particularly homicide, and whether this is affected by pandemic conditions.

@@

```{r, echo=FALSE}
# using latitude and longitude to obtain the zip codes of the crime locations
# crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))
# id_lat_long_crime_la <- subset(crime_la, select = c("LAT", "LON", "DR_NO", "year"))

@@

Based on this plot, we do not observe a clear trend between the number of crimes committed and the median income in that zip code. However, this scatterplot only accounts for a small portion of the dataset and averages across pre- and post-pandemic conditions. Future visualizations and analyses could consider pandemic vs non-pandemic years as variables.

# Exploratory Data Analysis With ggplot2

## Number of Evictions and Crime Levels Per Month
```{r, echo=FALSE}
# fix mistake
evictions_la$Date.Received <- gsub("2205", "2025", evictions_la$Date.Received)

# convert to date
evictions_la$month_year <- format(as.Date(evictions_la$Date.Received, format="%m-%d-%Y"), "%Y-%m")

# get unique months
unique_months <- data.frame(month_year = sort(unique(evictions_la$month_year)), index = 1:length(unique(evictions_la$month_year)))

evictions_la_indexed <- evictions_la %>% left_join(unique_months, by = "month_year")

# create color scheme
crime_la$month_year <- format(crime_la$DateParsed, "%Y-%m")
crime_la_monthly <- crime_la %>% 
  group_by(month_year) %>% 
  summarize(num_crimes = n())

evictions_la_indexed_crimes <- evictions_la_indexed %>% left_join(crime_la_monthly, by = "month_year")

avg_crimes <- mean(evictions_la_indexed_crimes$num_crimes)

evictions_la_indexed_crimes$high_low_crime <- as.factor(ifelse(evictions_la_indexed_crimes$num_crimes > avg_crimes, "Above Average", "Below Average"))

ggplot(data = evictions_la_indexed_crimes, aes(x = index)) + 
  geom_histogram(aes(fill = high_low_crime), bins = nrow(unique_months)) +
  scale_x_continuous(breaks=unique_months$index,
        labels=unique_months$month_year) + 
  labs(x = "Month and Year of Eviction Notice Receipt", y = "Number of Evictions", title = "Monthly Frequency of Evictions and Crimes in the City of Los Angeles \nFebruary 2023 to February 2025", fill = "Crime Level") +
  ggthemes::scale_fill_fivethirtyeight() + 
  ggthemes::theme_stata(scheme = "s1mono") +
  theme(axis.text.x = element_text(angle = 300, vjust = 0.5, hjust = -.1),
        axis.title.x = element_text(vjust = -.75),
        plot.title = element_text(size=12))
```
### Interpretation
This histogram plots the frequency of evictions in Los Angeles from February 2023 to February 2025, where the City of Los Angeles Housing Department has published case data. By summarizing both the crime and the evictions datasets by month, we see that the number of evictions was highest in early-mid 2023, and since, there have typically been slightly lower frequencies of evictions. Additionally, we split the number of crimes per month based on the average number of crimes of 14,439 to denote above and below average number of crimes per month. There is a clear split between March and April of 2024, as all months before that had above average crime, and all after had below average. Interestingly, the months where there is below average crime, seem to be the ones with slightly lower numbers of evictions.

## Boxplot - Kaitlyn

```{r, echo=FALSE}
crime_by_area <- crime_la %>% group_by(AREA.NAME) %>% summarize(numberOfCrimes = n()) %>% ungroup()

bureaus <- read.csv("Raw_Data/Division_Bureau_Ref.csv")

bureaus$Division <- gsub("77th", "77th Street", bureaus$Division)
bureaus$Division <- gsub("North", "N", bureaus$Division)
bureaus$Division <- gsub("Neast", "Northeast", bureaus$Division)

crime_by_area <- crime_by_area %>% left_join(bureaus, join_by("AREA.NAME" == "Division"))

ggplot(crime_by_area, aes(y = numberOfCrimes, group = Bureau, x = Bureau)) + 
        geom_boxplot(aes(fill = Bureau))
```

## geom_tile for correlations - Bianca

```{r, echo=FALSE}
library(lubridate)
# creating dataset for correlations

# since we do not have pre- and post- pandemic city of LA data, we are instead using LA county data for evictions. 
LA_county_evictions <- read.csv("Raw_Data/LA_county_evictions.csv")

# covid data
covid_14day$date <- ymd(covid_la$ep_date)
covid_14day$month <- month(covid_la$date)
covid_14day$year <- year(covid_la$date)
covid_14day$date <- day(covid_la$date)
# since there is a 7 day lag in reporting, let's do monthly = cases on the 21st and on the 5th of the next month.
covid_14day$lagged <- covid_la$date + 7

covid_la_lagged <- covid_14day %>% filter(lagged %in% c(21, 35)) %>% arrange(month, year)

covid_la_monthly <- covid_la_lagged %>% group_by(month, year) %>% summarize(cases = sum(cases_14day)) %>% ungroup() %>% arrange(year, month)

covid_la_monthly$index <- 1:nrow(covid_la_monthly)
# base plot shows that this does in fact mirror LA county so the case numbers should be accurate
# plot(y = covid_la_monthly$cases, x = covid_la_monthly$index, type = "l")

crime_la_monthly$date <- ymd(paste0(crime_la_monthly$month_year, "-01"))
crime_la_monthly$month <- month(crime_la_monthly$date)
crime_la_monthly$year <- year(crime_la_monthly$date)
```

## geom_jitter

## Plot 5

## Plot 6

@@

## Regression to Examine Weekly Crime Rates

```{r, echo=FALSE}
library(broom)

weekly_counts$CrimesLag <- c(NA, weekly_counts$Crimes[-nrow(weekly_counts)])

@@

ar1_trend_regression %>%
tidy() %>%
  mutate(
    p.value = scales::pvalue(p.value)
  ) %>%
  knitr::kable(
    caption = "Linear Regression Model Results Estimating Crimes Per Week",
    col.names = c("Predictor", "B", "SE", "t", "p"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "r", "r", "r", "r")
  )
```

@@

## Regression Between Median Income and Crimes per Zip Code

```{r, echo=FALSE}
lm_income_crimes <- lm(number_of_crimes ~ cali_median_income, data = crime_income_per_zip)

# summary(lm_income_crimes)

lm_income_crimes %>%
tidy() %>%
  mutate(
    p.value = scales::pvalue(p.value)
  ) %>%
  knitr::kable(
    caption = "Linear Regression Model Results Estimating Crimes Based on Median Income",
    col.names = c("Predictor", "B", "SE", "t", "p"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "r", "r", "r", "r")
  )
```

Based on the p-value, we observe that median income is not significant at a 0.05 level, meaning it does not significantly decrease the amount of crimes committed. The r\^2 value additionally tells us that the income only explains a small portion of the variation in the number of crimes. As stated with the scatterplot, a larger sample and considering other factors might yield different results.

# References {.unnumbered}

::: {#refs}
:::

# Data Sources {.unnumbered}
## Primary Datasets {.unnumbered}
Crime Data From 2010-2019: https://data.lacity.org/Public-Safety/Crime-Data-from-2010-to-2019/63jg-8b9z/about_data

Crime Data From 2020-Present: https://catalog.data.gov/dataset/crime-data-from-2020-to-present

## Secondary Datasets {.unnumbered}
Los Angeles County Eviction Data: https://calmatters.org/housing/homelessness/2023/11/california-evictions-post-pandemic/

City of Los Angeles Eviction Data: https://housing.lacity.gov/residents/renters/eviction-notices-filed

Median Income Data: https://data.census.gov/table/ACSST1Y2023.S1903?q=income

# Appendices {.appendix}

## Data Wrangling {.unnumbered}

# Reading in Raw Datasets

```{r Wrangling Main, eval=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses

# Combine the two datasets
crime_la <- rbind(crime_la_2010_2019, crime_la_2020_present)
```

```{r Combining Median Income, eval=FALSE}
# Secondary datasets
california_median_income_2018_2023 <- list()
yrs <- 2018:2023
for (i in 1:length(yrs)) {
  obj <- read.csv(paste0("Raw_Data/ACSST5Y",yrs[[i]],".S1903-Data.csv"), skip = 0)
  obj <- obj[-1,]
  obj$year <- yrs[[i]]
  california_median_income_2018_2023[[i]] <- obj
}

# adding a variable that's just the ZCTA zip code
cali_median_income$ZCTA <- stringr::str_sub(cali_median_income$NAME, -5)

write.csv(cali_median_income, "cali_median_income.csv")
```

### Zip Codes for Crimes {.unnumbered}
```{r Get Zip Codes, eval=FALSE}
library(sf)
library(dplyr)

crime_la_2010_2019 <- read.csv("Raw_Data/Crime_Data_from_2010_to_2019.csv")

crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))

coords <- crime_la %>% select(LAT, LON, DR_NO, year)

zcta <- read_sf(dsn = "Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")

  points <- st_as_sf(coords, coords = c("LON", "LAT"), crs = 4326)  # WGS84

  zcta <- st_transform(zcta, 4326)  # Convert ZCTA to WGS84 if needed
  cat("Starting Zip Code Retrieval \n")

  result <- st_join(points, zcta)

  cat("Finished Zip Code Retrieval \n")

  write.csv(result, "crime_zip_files/all_zips.csv")

crime_zctas <- read.csv("Raw_Data/all_zips.csv")
colnames(crime_zctas) <- c(colnames(crime_zctas)[-c(1, length(colnames(crime_zctas)))],"LON", "LAT")

crime_zctas$LON <- as.numeric(gsub('c(', "", crime_zctas$LON, fixed = TRUE))
crime_zctas$LAT <- as.numeric(gsub(')', "", crime_zctas$LAT, fixed = TRUE))
```

```{r Get Zip Codes 2, eval=FALSE}

duplicates <- as.integer(as.character(data.frame(table(crime_zctas$universal_idx)[table(crime_zctas$universal_idx) > 1]) %>% pull(Var1)))

# there are approximately 200 rows that are duplicated (100 incidents) as the area lies on the intersection between two zctas. 
duplicate_data <- crime_zctas %>% filter(universal_idx %in% duplicates)

rev_geo_duplicates <- reverse_geocode(duplicates2, lat = LAT, long = LON)
# the above query identifies the address of these coordinates as KFC, 8644, Balboa Boulevard, Northridge South Neighborhood Council District, Los Angeles, Los Angeles County, California, 91325, United States. 
# therefore, we will only keep the 91325 zip code rows.

crime_zctas_no_duplicates <- crime_zctas %>% mutate(duplicate = ifelse(universal_idx %in% duplicate_data$universal_idx & ZCTA5CE20 != 91325, "Drop", "Keep")) %>% filter(duplicate != "Drop") %>% select(-duplicate)

# it's empty now 
# duplicates <- table(crime_zctas_no_duplicates$universal_idx)[table(crime_zctas_no_duplicates$universal_idx) > 1]

# next, check for NAs
sum(is.na(crime_zctas_no_duplicates$ZCTA5CE20)) # there are 3724 NAs. Let's find them a zip code

# some also have lat long = 0. We'll find those separately
no_zipcode_not0 <- crime_zctas_no_duplicates %>% filter(is.na(ZCTA5CE20) & LON != 0)

no_zipcode_not0_zips <- reverse_geocode(no_zipcode_not0, lat = LAT, long = LON)

no_zipcode_not0_zips$ZCTA5CE20 <- stringr::str_extract(no_zipcode_not0_zips$address, "\\d{5}")

no_zipcode_not0_zips <- no_zipcode_not0_zips %>% select(-address)

crime_zctas2 <- crime_zctas_no_duplicates %>% filter(!is.na(ZCTA5CE20))

crime_zctas3 <- rbind(crime_zctas2, no_zipcode_not0_zips)

crime_zctas4 <- crime_zctas3 %>% filter(!is.na(ZCTA5CE20))

write.csv(crime_zctas4, "crimes_with_zips.csv")

no_zipcode_0s <- crime_zctas_no_duplicates %>% filter(is.na(ZCTA5CE20) & LON == 0) %>% pull(DR_NO)

main_dataset_0s <- crime_la %>% filter(DR_NO %in% no_zipcode_0s)

summary(main_dataset_0s)

summary(crime_la)
```

# Figure 1 - Examining Crime Density from 2010–2019 and 2020-Present {.unnumbered}

```{r Figure 1 Code, eval=FALSE}

# Figure 2 - Examining Number Of Crimes Over Time {.unnumbered}

```{r Figure 2 Code, eval=FALSE}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

# Figure 3 - Crime rate frequency by day of the week {.unnumbered}

```{r Figure 3 Code, eval=FALSE}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

# Plot bar chart with horizontal labels
barplot(crime_counts, col = "lightblue",
        xlab = "Day of the Week", ylab = "Number of Crimes",
        names.arg = names(crime_counts), las = 1, cex.names = 0.8, cex.axis = 0.6)
```

# Figure 4 - Gender Distribution for crimes over time {.unnumbered}

```{r Figure 4 Code, eval=FALSE}
# Remove missing values and empty strings in Vict.Sex column
crime_la <- subset(crime_la, Vict.Sex != "" & !is.na(Vict.Sex))

# Figure 5 - Median Household Income by Zip Code in Los Angeles by Year {.unnumbered}

```{r Figure 5 Wrangling, eval=FALSE}
zcta <- read_sf(dsn = "Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
LA_zctas <- read.csv("Raw_Data/zctas.csv") #obtained using zipcodeR::search_city("Los Angeles", "CA") and then checking which zip codes are also ZCTAs

```{r Figure 5 Code, eval=FALSE}
# quantiles for the entire dataset (from 2018 to 2023) to divide the median income into groups
qntls <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

# Figure 6 - Crime Rates Compared to Median Household Income {.unnumbered}

```{r Figure 6 Code, eval=FALSE}
# using latitude and longitude to obtain the zip codes of the crime locations
# crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))
# id_lat_long_crime_la <- subset(crime_la, select = c("LAT", "LON", "DR_NO", "year"))

## Crime \~ Week Regression Code {.unnumbered}

```{r Regression 1, eval=FALSE}
library(broom)

weekly_counts$CrimesLag <- c(NA, weekly_counts$Crimes[-nrow(weekly_counts)])

ar1_trend_regression %>%
tidy() %>%
  mutate(
    p.value = scales::pvalue(p.value)
  ) %>%
  knitr::kable(
    caption = "Linear Regression Model Results Estimating Crimes Per Week",
    col.names = c("Predictor", "B", "SE", "t", "p"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "r", "r", "r", "r")
  )
```

## Crime \~ Income Regression Code {.unnumbered}

```{r Regression 2, eval=FALSE}
lm_income_crimes <- lm(number_of_crimes ~ cali_median_income, data = crime_income_per_zip)

summary(lm_income_crimes)

lm_income_crimes %>%
tidy() %>%
  mutate(
    p.value = scales::pvalue(p.value)
  ) %>%
  knitr::kable(
    caption = "Linear Regression Model Results Estimating Crimes Based on Median Income",
    col.names = c("Predictor", "B", "SE", "t", "p"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "r", "r", "r", "r")
  )
```

>>>>

author: "Marco Stine, Kevin Hu, Bianca Schutz, and Kaitlyn Kirt"
fontsize: 10pt
toc: true                   # add table of contents at the beginning
toc-depth: 3               # Only titles that start with # or ##
geometry: margin=1in
bibliography: references.bib
format:
  pdf: default
---

@@

# Loading in packages and dataset

```{r, output=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses

@@

# Secondary datasets
california_median_income_2018_2023 <- list()
yrs <- 2018:2023
for (i in 1:length(yrs)) {
  obj <- read.csv(paste0("Raw_Data/ACSST5Y",yrs[[i]],".S1903-Data.csv"), skip = 0)
  obj <- obj[-1,]
  obj$year <- yrs[[i]]
  california_median_income_2018_2023[[i]] <- obj
}

# adding a variable that's just the ZCTA zip code
cali_median_income$ZCTA <- stringr::str_sub(cali_median_income$NAME, -5)
```

# Plots

@@

```{r}

@@

```{r}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

@@

@@

```{r}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

# Plot bar chart with horizontal labels
barplot(crime_counts, col = "lightblue",
        main = "Crime Occurrences by Day of the Week",
        xlab = "Day of the Week", ylab = "Number of Crimes",
        names.arg = names(crime_counts), las = 1, cex.names = 0.8, cex.axis = 0.6)
```

@@

```{r}
# Remove missing values and empty strings in Vict.Sex column
crime_la <- subset(crime_la, Vict.Sex != "" & !is.na(Vict.Sex))

@@

```{r}
zcta <- read_sf(dsn = "Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
LA_zctas <- read.csv("Raw_Data/zctas.csv") #obtained using zipcodeR::search_city("Los Angeles", "CA") and then checking which zip codes are also ZCTAs

@@

Now, using this joined dataset, create six maps, one for each year.

```{r}
# quantiles for the entire dataset (from 2018 to 2023) to divide the median income into groups
qntls <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

@@

### Interpretation
The maps demonstrate that median income in certain LA zip codes (particularly those to the west) have a much higher median household income than the rest of the city, but over the last six years, the median income in most other LA zip codes has increased. However, there are some that a typical family is at or below the poverty line. We intend to investigate if there is a spatial correlation between these poorer zip codes and the incidence of crimes, particularly homicide, and whether this is affected by pandemic conditions.

@@

```{r}
# using latitude and longitude to obtain the zip codes of the crime locations
# crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))
# id_lat_long_crime_la <- subset(crime_la, select = c("LAT", "LON", "DR_NO", "year"))

@@

### Based on this plot, we do not observe a clear trend between the number of crimes committed and the median income in that zip code. However, this scatterplot only accounts for a small portion of the dataset and averages across pre- and post-pandemic conditions. Future visualizations and analyses could consider pandemic vs non-pandemic years as variables.

@@

## Using Regression to Examine Weekly Crime Rates

```{r}
weekly_counts$CrimesLag <- c(NA, weekly_counts$Crimes[-nrow(weekly_counts)])

@@

summary(ar1_trend_regression)
```
### Interpretation

@@

## Basic Regression Between Median Income and Crimes per Zip Code

```{r}
lm_income_crimes <- lm(number_of_crimes ~ cali_median_income, data = crime_income_per_zip)

summary(lm_income_crimes)
```
### Interpretation

Based on the p-value, we observe that median income is not significant at a 0.05 level, meaning it does not significantly decrease the amount of crimes committed. The r^2 value additionally tells us that the income only explains a small portion of the variation in the number of crimes. As stated with the scatterplot, a larger sample and considering other factors might yield different results.
